---
title: "Global Economic Indicators: Trend Analysis (2005–2025)"
author: "LaToya Burdette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

library(tidyverse)
library(scales)
library(patchwork)

theme_set(
  theme_light(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      plot.caption = element_text(size = 9)
    )
)
```

## Overview

This report looks at global economic indicators from 2005–2025, including GDP, GDP growth, inflation, and unemployment. Major economic events are included to help explain changes in these trends over time.

```{r}
world <- read_csv("data/global_economy_world_aggregate.csv", show_col_types = FALSE) %>%
  mutate(
    year = as.integer(year),
    global_gdp_per_capita = global_gdp_usd / global_population
  )

events <- read_csv("data/economic_events_timeline.csv", show_col_types = FALSE) %>%
  mutate(year = as.integer(year))

```

## Global GDP (USD)

```{r}
ggplot(world, aes(x = year, y = global_gdp_usd)) +
  geom_line(linewidth = 1.2, color = "#2C7FB8") +
  scale_x_continuous(breaks = seq(min(world$year), max(world$year), by = 5)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Global GDP Expansion (2005–2025)",
    subtitle = "Long-term upward trend with periods of disruption",
    x = "Year",
    y = "Global GDP (USD)"
  )
```

Global GDP generally trends upward over time, although there are clear slowdowns and declines during major global disruptions.

## Global GDP Growth (%) with Event Markers

```{r}
ggplot(world, aes(x = year, y = global_gdp_growth_pct)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(linewidth = 1.2, color = "#D95F02") +
  geom_vline(data = events, aes(xintercept = year), linetype = "dashed", alpha = 0.4) +
  labs(
    title = "Global GDP Growth (%)",
    x = "Year",
    y = "Growth (%)"
  )
```
Large economic events often correspond with changes in the direction or pace of global growth.

## Global Inflation (%)

```{r}
ggplot(world, aes(x = year, y = global_inflation_pct)) +
  geom_line(linewidth = 1.2, color = "#1B9E77")+
  scale_x_continuous(breaks = seq(min(world$year), max(world$year), by = 5)) +
  labs(
    title = "Global Inflation (%)",
    x = "Year",
    y = "Inflation (%)"
  )
```

## Global Unemployment (%)

```{r}
ggplot(world, aes(x = year, y = global_unemployment_pct)) +
  geom_line(linewidth = 1.2, color = "#7570B3") +
  scale_x_continuous(breaks = seq(min(world$year), max(world$year), by = 5)) +
  labs(
    title = "Global Unemployment (%)",
    x = "Year",
    y = "Unemployment (%)"
  )
```

## GDP Per Capita

```{r}
ggplot(world, aes(x = year, y = global_gdp_per_capita)) +
  geom_line(linewidth = 1.2, color = "#E7298A") +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
  scale_x_continuous(breaks = seq(min(world$year), max(world$year), by = 5)) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Global GDP Per Capita",
    subtitle = "The dashed line shows the overall long-term trend",
    x = "Year",
    y = "USD"
  )
``` 
The dashed line highlights the overall long-term trend.

## Summary Statistics

```{r}
summary_stats <- world %>%
  summarise(
    avg_growth = mean(global_gdp_growth_pct, na.rm = TRUE),
    sd_growth = sd(global_gdp_growth_pct, na.rm = TRUE),
    avg_inflation = mean(global_inflation_pct, na.rm = TRUE),
    avg_unemployment = mean(global_unemployment_pct, na.rm = TRUE)
  )

knitr::kable(summary_stats, digits = 2)
```

## Conclusion

 Global economic trends from 2005–2025 show steady long-term growth with periods of volatility. Major economic disruptions are reflected in changes to GDP growth, inflation, and unemployment patterns.
 
 